---
title: "Index of housing retrofit need"
author: "Prof Ben Barr, Dr Eleojo Abubakar et al."
date: "2025-10-03"
output: html_document
---
# Introduction

Aim: Create an index at the national level that captures how neighbourhood-level needs for housing retrofit vary in a way that is sensitive to differences in population health status. 

Question: To what extent do neighbourhood-level needs for housing retrofits differ according to housing energy efficiency, levels of income deprivation, and the vulnerability of residents to poor housing conditions due to their health status? And how does this vary across regions of England? 

# Methods summary
Neighbourhood retrofit need was operationalised at the Lower-layer Super Output Area (LSOA) level for England by combining three domains: housing energy inefficiency, income deprivation, and population health vulnerability. Housing energy inefficiency was measured as the percentage of dwellings with an Energy Performance Certificate (EPC) rating of band D or worse, derived from NOMIS small-area housing energy efficiency statistics. Income deprivation was captured using the Income domain score/rate from the 2019 Index of Multiple Deprivation (IMD). Health vulnerability indicators were assembled from two sources: (i) Quality and Outcomes Framework (QOF) prevalence estimates for selected long-term conditions (including asthma, chronic obstructive pulmonary disease (COPD) and mental health) and (ii) LSOA-level emergency admission rate indicators for condition groups plausibly sensitive to poor housing (including respiratory conditions, cardiovascular disease, and injuries/burns/falls among adults aged ≥65 years). Datasets were harmonised to a common LSOA geography using best-fit lookup tables (including LSOA 2011–2021 correspondences where required), and analyses were restricted to English LSOAs.

A composite retrofit needs index was generated using a multi-criteria decision analysis approach. Missing values in numeric indicators were imputed using the corresponding local-authority (district) mean to preserve small-area coverage, and extreme values were reduced using percentile-based winsorisation. COPD and asthma prevalence were additionally combined using a lightly trimmed mean to stabilise estimates while retaining the original percentage scale. Health vulnerability indicators (all expressed as percentages/rates on a comparable scale) were aggregated into a single health-vulnerability domain using a weighted mean, with condition weights informed by published estimates of health impacts associated with meeting housing quality standards (Rodgers et al., 2018) and normalised to sum to one before application. The final retrofit needs index combined the three domains (energy inefficiency, income deprivation, and health vulnerability) using a weighted-product formulation to reflect compounding need when multiple domains are jointly high. Index values were then standardised (z-scores), expressed as a relative index (x/mean(x)), ranked, and mapped using LSOA boundary shapefiles in R, with local authority boundaries overlaid for geographic interpretation.

# Environment setup and data import
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(max.print = 10000, timeout = max(600, getOption("timeout")))

library(readxl, quietly = T)
library(SmartEDA, quietly = T)
library(tidyverse, verbose = F)
library(magrittr)

# plots
library(scales)
library(patchwork)
library(plotly)
library(ggrepel)
library(ggsankeyfier)
library(paletteer)
library(ragg)

#for data preprocessing
library(sjmisc)
library(janitor)

library(statsExpressions)
library(lares) #compute correlations

# for mapping
library(sf)
library(tmap)
tmap_options(
  output.dpi = 600,
  unit = "cm"
)

#make 'outputs' directory (if it doesn't already exist)
outputs_dir <- "outputs"
if(!dir.exists(outputs_dir)){
  dir.create(outputs_dir)
}

# The GitHub-facing workflow is the default. Set
# HOUSING_RETROFIT_DATA_MODE=safeguarded only in the approved secure environment
# to rebuild the index from the restricted emergency-admissions inputs.
run_safeguarded <- identical(
  tolower(Sys.getenv("HOUSING_RETROFIT_DATA_MODE", unset = "public")),
  "safeguarded"
)

public_index_path <- file.path(
  "open_data",
  "housing_retrofit_need_index_lsoa_v3.csv"
)

# Stable publisher endpoints. Large public files are downloaded to the R
# session's temporary directory rather than committed to this repository.
source_urls <- list(
  nomis_energy = paste0(
    "https://www.nomisweb.co.uk/api/v01/dataset/NM_2401_1.data.csv?",
    "date=2024&geography=TYPE151&energy_efficiency_8=4...7&",
    "measures=20100,20301"
  ),
  imd2019 = paste0(
    "https://assets.publishing.service.gov.uk/media/",
    "5d8b3b51ed915d036a455aa6/File_5_-_IoD2019_Scores.xlsx"
  ),
  qof_asthma = "https://pldr.org/download/e6nzv/ng1/QOF_4_03_Asthma_LSOA.csv",
  qof_copd = "https://pldr.org/download/23q1e/2tm/QOF_4_04_COPD_LSOA.csv",
  qof_mental_health = "https://pldr.org/download/ex3od/j5c/QOF_4_15_MentalHealth_LSOA.csv",
  lsoa_2011_to_2021 = paste0(
    "https://open-geography-portalx-ons.hub.arcgis.com/api/download/v1/items/",
    "b684a0dbf786473f9563ec0616da2f8b/csv?layers=0"
  ),
  lsoa_2011_to_lad_2020 = paste0(
    "https://open-geography-portalx-ons.hub.arcgis.com/api/download/v1/items/",
    "31dfd51115194a6bbe0bf4a26019f884/csv?layers=0"
  ),
  lsoa_2011_boundaries = paste0(
    "https://open-geography-portalx-ons.hub.arcgis.com/api/download/v1/items/",
    "f23b8af6504640558a5100dfcd19a7ee/geojson?layers=0"
  ),
  lad_2011_boundaries = paste0(
    "https://open-geography-portalx-ons.hub.arcgis.com/api/download/v1/items/",
    "ad7f01a8ae73441b95444080c78caa17/geojson?layers=0"
  ),
  lsoa_2011_to_region = paste0(
    "https://open-geography-portalx-ons.hub.arcgis.com/api/download/v1/items/",
    "c1e13af610c84ec59f086502e8ebe4f7/csv?layers=0"
  )
)

download_to_temp <- function(url, fileext) {
  destination <- tempfile(fileext = fileext)
  download.file(url, destination, mode = "wb", quiet = TRUE)
  destination
}

# data prep for shiny app
# tmp_fld <- "E:/UniLiverpool/data_analysis/phirst_housing_retrofit/retrofit_shiny_app/retrofit_shiny_app/data/default"
# if(!dir.exists(tmp_fld)){
#   dir.create(tmp_fld)
# }
```


```{r data-mode}
# Secure paths are supplied through environment variables and are never stored
# in the public repository.
if (run_safeguarded) {
  emgcy_adm_root_v2 <- Sys.getenv("EMERGENCY_ADMISSIONS_ROOT")
  lsoa_pop_str_root <- Sys.getenv("LSOA_POPULATION_ROOT")

  missing_secure_paths <- c(
    EMERGENCY_ADMISSIONS_ROOT = emgcy_adm_root_v2,
    LSOA_POPULATION_ROOT = lsoa_pop_str_root
  )
  missing_secure_paths <- names(missing_secure_paths)[
    !dir.exists(unname(missing_secure_paths))
  ]

  if (length(missing_secure_paths) > 0) {
    stop(
      "Safeguarded mode requires valid directories in: ",
      paste(missing_secure_paths, collapse = ", ")
    )
  }
}

```

# House energy Conditions
Download data from <https://www.nomisweb.co.uk/query/construct/submit.asp?menuopt=201&subcomp=>

1. House conditions: % of households with EPC <= d
see the column "epc_lss_eql_d_prcnt"
```{r}
# import energy efficiency data
engy_eff <- read_csv("./open_data/housing_energy_efficiency_nomis.csv", skip = 5) %>% 
  clean_names() %>% 
  glimpse

# rename columns
# compute % less than or equal to EPC band D
engy_eff <- 
engy_eff %>% 
  rename(., any_of(setNames(c("percent_4", "percent_6", "percent_8", "percent_10", "percent_12", "percent_14", "percent_16", "percent_18"), c("total_prcnt", "band_a_prcnt", "band_b_prcnt", "band_c_prcnt", "band_d_prcnt", "band_e_prcnt", "band_f_prcnt", "band_g_prcnt")))
  ) %>% 
  rowwise() %>% 
  mutate(
    # compute % of households with EPC <= d
    epc_lss_eql_d_prcnt = sum(band_d_prcnt, band_e_prcnt, band_f_prcnt, band_g_prcnt),
    epc_lss_eql_d_no = sum(band_d, band_e, band_f, band_g)
  ) %>% 
  ungroup() %>% 
  glimpse

```

# Low income
Income score of IMD 2019 or income deprivation affecting older people / income deprivation affecting children.

see the column named "income_score_rate" or "income_deprivation_affecting_children_index_idaci_score_rate" or "income_deprivation_affecting_older_people_idaopi_score_rate"
```{r imd2019}
# import IMD2019 score
imd2019 <- read_excel(
  download_to_temp(source_urls$imd2019, ".xlsx"),
  sheet = 2
  ) %>% 
  clean_names() %>% 
  glimpse

imd2019 %>% glimpse

```

# Population structure
% over 65

Can download open LSOA data of 2011 - 2022 from <https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/lowersuperoutputareamidyearpopulationestimates>
see associated documentation at <https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/methodologies/smallareapopulationestimatesqmi>

but use alternative safeguarded data available at ILRR_Store/Safeguarded_Data/ELF_LSOA_data, for 2014 - 2022
```{r population-structure, eval=run_safeguarded}
## Get full file paths of all CSVs (matching the specified regex pattern)
lsoa_pop_65plus <- list.files(path = lsoa_pop_str_root, pattern = "^LSOA_65plus_mod_.+\\.csv$", full.names = TRUE)
lsoa_pop_65plus

# if prefer data not modelled, use the following regex instead
# lsoa_pop_65plus <- list.files(path = lsoa_pop_str_root, pattern = "^LSOA_65plus_sev_.+\\.csv$", full.names = TRUE)

## Create clean names (without prefix and extension)
df_names2 <- str_remove_all(basename(lsoa_pop_65plus), "^LSOA_65plus_|\\.csv$")
df_names2

## Ingest all CSV files in folder into a list of named dfs
pop65plus_lsoa_mod <- lsoa_pop_65plus %>%
  set_names(df_names2) %>%
  map(read_csv)

# explore data
names(pop65plus_lsoa_mod)
pop65plus_lsoa_mod[["mod_sev_2022"]] %>% glimpse

```

# Prevalence of health condition in LSOA

Update: 03/08/2026
Use PLDR published LSOA prevalence data

The R Markdown file reads `QOF_4_04_COPD`, `QOF_4_03_Asthma`, and `QOF_4_15_MentalHealth` directly from PLDR, replacing the original bespoke script for creating the QOF prevalence indicators.
QOF_4_04: <https://pldr.org/dataset/quality-and-outcomes-framework-indicators-chronic-obstructive-pu-23q1e>
QOF_4_03: <https://pldr.org/dataset/quality-and-outcomes-framework-indicators-asthma-prevalence-qof4-e6nzv>
QOF_4_15 (Mental health): <https://pldr.org/dataset/quality-and-outcomes-framework-mental-health-prevalence-qof415-ex3od>

```{r qof-prevalence, eval=run_safeguarded}
# Read the three published PLDR CSVs directly. They are deliberately not
# duplicated in open_data because each source file is approximately 40 MB.
qof_prev_urls <- c(
  QOF_4_03_Asthma = source_urls$qof_asthma,
  QOF_4_04_COPD = source_urls$qof_copd,
  QOF_4_15_MentalHealth = source_urls$qof_mental_health
)

prev_lsoa_lst <- map(
  qof_prev_urls,
  ~ read_csv(.x, show_col_types = FALSE)
)

# prev_lsoa_lst[[1]] %>% 
#   # frq(year)
#   glimpse

# slice and cast pertinent LSOA prevalence data
hlth_cndns_prev_lsoa <-
  prev_lsoa_lst %>%
  bind_rows(.id = "health_condition") %>% 
  filter(year == 2022) %>%
  # select(-c(den, num)) %>% 
  pivot_wider(
    id_cols = c(lsoa11, year),
    names_from = health_condition,
    names_glue = "{health_condition}.{.value}",
    values_from = c(p_rate, den, num),
    # names_sep = "."
  ) %>% 
  rename_with(~str_split_i(.x, regex("\\d{2}_"), -1), .cols = starts_with("QOF_")) %>% 
    rowwise() %>% 
  mutate(
    # mean prev of asthma and copd rates
    copd_asthma.p_rate = mean(c(COPD.p_rate, Asthma.p_rate), trim = 0.025, na.rm = T)
  ) %>% 
    ungroup() %>% 
  glimpse

# hlth_cndns_prev_lsoa %>% 
#   glimpse

```

# Disease severity Data
This is data on emergency hospital admission at the LSOA level for 2007 to 2019, safeguarded in the university

Update: Rate computed here for emergency hospitalisation is now multiplied by 100, like other QoF prevalence rates.
```{r disease-severity, eval=run_safeguarded}
#-------------------------------------
# DISEASE SEVERITY DATA (VERSION 2)
#-----------------------------------
# APC_2_03: All emergency admissions
# APC_2_05: Asthma
# APC_2_07: Respiratory conditions
# APC_2_24: CVD
# APC_2_49: Burns/Falls related hospital admissions
# -----------------------------

## Get full file paths of all CSVs (matching the specified regex pattern)
## for burns/falls, get file disaggregated by age
emgcy_adm_files <- list.files(path = emgcy_adm_root_v2,
                              pattern = "^PLDR_v5_2_.+_APC_all_byLSOA\\.csv$|.+v5_2_49_APC_age5_byLSOA+.",
                              full.names = TRUE,
                              recursive = T
                              )
emgcy_adm_files

# exclude file representing all conditions from list and
# exclude file representing injuries and falls, not disaggregated by age
emgcy_adm_cnd_files <- emgcy_adm_files[2:(length(emgcy_adm_files)-1)]
emgcy_adm_cnd_files

# read all files, process and stack them
for(x in seq_along(emgcy_adm_cnd_files)){
  if(x == 1){
    emgcy_adm_df_lst = list()
    cndns = c("asthma", "respiratory", "cvd", "burns_falls_65_ova")
  }
  
  # read files to named list of dataframes
  emgcy_adm_df_lst[[cndns[x]]] <- read_csv(emgcy_adm_cnd_files[x],
                                           show_col_types = FALSE)
  
  # if agegroup in column names,
  # recompute prevalence rate (p_rate) for aggreg age 65+ as p_num/mid_pop
  if("agegroup" %in% colnames(emgcy_adm_df_lst[[cndns[x]]])){
    print("processing falls/injuries for age 65+ ...")
    emgcy_adm_df_lst[[cndns[x]]] <- emgcy_adm_df_lst[[cndns[x]]] %>% 
      filter(agegroup >= 65) %>% 
      group_by(lsoa11, year) %>% 
      summarise(
        p_num = sum(p_num),
        admi_days = sum(admi_days),
        non_na = sum(non_na),
        mid_pop = sum(mid_pop)
      ) %>% 
      ungroup() %>% 
      mutate(
        mid_pop = case_when(
          mid_pop <= 0 ~ NA,
          p_num/mid_pop > 0.3 ~ NA,
          .default = mid_pop
        ),
        p_rate=(p_num/mid_pop)*100
      )
  }
  
  # after ingesting all files, stack!
  if(x == length(emgcy_adm_cnd_files)){
    emgcy_adm_df_v2 <- bind_rows(emgcy_adm_df_lst, .id = "hlth_cndn")
    emgcy_adm_df_v2 %>% glimpse
  }
}

# explore dataframe v2
emgcy_adm_df_v2 %>% glimpse
  frq("hlth_cndn", "year")

# cast data (i.e. long to wide format) for onward processing
emgcy_adm_df_v2_cst <- 
emgcy_adm_df_v2 %>% 
  mutate(
    name = hlth_cndn
         ) %>% 
  pivot_wider(
    id_cols = c(lsoa11, year),
    names_from = name,
    names_glue = "{name}.{.value}",
    values_from = p_rate, #p_num and/or mid_pop can be added here if needed
  ) %>% 
  filter(year == 2022) %>% 
  glimpse

# save project data
gc()
save.image()

```

# Compute Index of Need for housing retrofit

We might have an issue of temporal mis-match since not all data sources have records upto 2024 to match EPC, LSOA prevalence etc data. Eg, latest age structure data is for 2022

## Load & normalise weights

These weights were derived from Rodgers et al (2018) and normalized to that all weights add up to one.

Rodgers SE, Bailey R, Johnson R, et al. Health impact, and economic value, of meeting housing quality standards: a retrospective longitudinal data linkage study. Public Health Research 2018;6(8):134.
```{r health-weights, eval=run_safeguarded}
# read table of weights
hlth_cnds_wts <- 
  read_xlsx("open_data/disease_wts_use.xlsx") %>% 
  filter(!is.na(label)) %>% 
  glimpse

# replace weight with average
hlth_cnds_wts[5,2] <- hlth_cnds_wts[5,3] #%>% 
hlth_cnds_wts #%>% glimpse

# confirm replacement
hlth_cnds_wts$wt
hlth_cnds_wts$...3

# save weights in a list with named values
for(i in seq_along(hlth_cnds_wts$label)){
  if(i == 1){
    wts = list()
  }
  
  wts[[hlth_cnds_wts$label[i]]] <- 
hlth_cnds_wts %>% 
  filter(label == hlth_cnds_wts$label[i]) %>% 
  pull(wt)
  
  # print(wts[i])
  
}

# flatten weights
wts_flt <- wts %>% unlist
wts_flt

# normalise weights so they all sum to one
wts_norm <- wts_flt/sum(wts_flt)
wts_norm

```

## Join pertinent LSOA data

Need to load data for weighting the LSOA measures

NB:
* Update to combination of copd + asthma rates. Now trimmed mean to maintain thesame scale as other rates (instead of weighted product).
* Year of QoF prevalence data now downgraded from 2024 to 2022, for better alignment with the latest data on emergency admissions.
* We have data for emergency admission due to asthma, and prevalence of asthma. We drop emergency admission due to asthma, and use only asthma prevalence rate (because we have weighting for it from Rodgers et al's 2018 paper)

```{r join-index-inputs, eval=run_safeguarded}
engy_eff %>% glimpse #based on LSOA 2021
imd2019 %>% glimpse #based on LSOA 2011
emgcy_adm_df_v2_cst %>% glimpse #2022 data based on LSOA 2011 (safeguarded)
pop65plus_lsoa_mod[["mod_sev_2022"]] %>% glimpse ##based on LSOA 2011
hlth_cnds_wts %>% glimpse
hlth_cndns_prev_lsoa %>% glimpse 

# LSOA 2011 to 2021 mapping
lsoa_11_to_21 <- 
  read_csv(source_urls$lsoa_2011_to_2021, show_col_types = FALSE) %>%
  select(
    -starts_with("LAD22"),
    -any_of(c("ObjectId", "ObjectID", "OBJECTID", "LSOA11NM"))
  ) %>% 
  glimpse

# lsoa_11_to_21 %>% glimpse

# data joins
## IMD inc + cond_prevelence + emergency admission + energy eff
retrofit_need_indx_lsoa <-
  imd2019 %>% select(LSOA11CD =lsoa_code_2011, income_score_rate) %>%
  left_join(hlth_cndns_prev_lsoa %>% select(lsoa11, MentalHealth.p_rate, copd_asthma.p_rate), by = c("LSOA11CD" = "lsoa11")) %>% 
  left_join(emgcy_adm_df_v2_cst %>% select(-c(year, asthma.p_rate)), by = c("LSOA11CD" = "lsoa11")) %>% 
  left_join(pop65plus_lsoa_mod[["mod_sev_2022"]] %>% select(lsoa11, prop_65_plus = r_mod_sev), by = c("LSOA11CD" = "lsoa11")) %>%
  left_join(lsoa_11_to_21, by = "LSOA11CD") %>% 
  left_join(engy_eff %>% select(LSOA21CD = mnemonic, epc_lss_eql_d_prcnt, epc_lss_eql_d_no)) %>%
  glimpse

retrofit_need_indx_lsoa %>% 
  glimpse

```

## Compute and export the index v3

A new workflow is proposed.

1. Fill missing values (in asthma.p_rate, 14 & prop_65_plus, 29) with their LA averages.

2. Apply 95% winsorization, all data below the 2.5th percentile set to the 2.5th percentile, and data above the 97.5th percentile set to the 97.5th percentile.

3. All indicators under the health domain are combined into a health vulnerability index, using weighted trimmed mean. All of these indicators are on thesame scale (%) and so do not need standardisation. Weighting was based on Rodgers et al, but all weights were normalised so they add up to one prior to application.

4. Both income and health vulnerability domains are on thesame scale (% individuals), but the EPC domain is on a different scale (% houses). So all three domains are combined using the weighted product method. 

5. Perform the weighted product method of all three domain

6. Standardise the resulting scores (by z_score & as a ratio of the grand mean).

The creation of a composite indicator in this study is a multicriteria decision analysis problem, we used both weighted sum and weighted product in this study.
See:
Mateo, J.R.S.C. (2012). Weighted Sum Method and Weighted Product Method. In: Multi Criteria Analysis in the Renewable Energy Industry. Green Energy and Technology. Springer, London. https://doi.org/10.1007/978-1-4471-2346-0_4

see also <https://www.oecd.org/content/dam/oecd/en/publications/reports/2008/08/handbook-on-constructing-composite-indicators-methodology-and-user-guide_g1gh9301/9789264043466-en.pdf>

```{r prepare-index-inputs, eval=run_safeguarded}

# 1. Fill missing values (in asthma.p_rate, 14 & prop_65_plus, 29) with their LA averages.

# LSOA to LAD mapping
LSOA11toLAD20 <- read_csv(
  source_urls$lsoa_2011_to_lad_2020,
  show_col_types = FALSE
  ) %>% 
  select(-any_of(c("FID", "ObjectId", "ObjectID", "OBJECTID"))) %>% 
  glimpse

# LSOA11toLAD20 contains 34,753 lsoas, while the lsoa21 contains 35,797 zones

rtrft_need_indtrs_fll <-
  retrofit_need_indx_lsoa %>% 
  left_join(LSOA11toLAD20, by = "LSOA11CD") %>% 
  select(-c(WD20CD, WD20NM)) %>%
  relocate(ends_with(c("NM", "CD"), ignore.case = F)) %>% 
    # get only LSOAs in England
  filter(str_starts(LSOA21CD, "E")) %>%
    relocate(LSOA21CD, LSOA21NM, .after = LSOA11NM) %>%
    # fill NAs with LAD means
    group_by(LAD20CD) %>% 
    mutate(
      across(
        where(is.numeric),
        ~ if_else(is.na(.x),
                if_else(all(is.na(.x)), NA_real_, mean(.x, na.rm = TRUE)),
                .x)
      )
    ) %>%
  ungroup() %>%
    glimpse

```

Two version of the raw composite index are created, (1) considers all 3 domains, the second considers only the energy inefficiency and income deprivation domains which is the popular approach in policy cycles. The essence is to uncover the sensitivity of the health-aware approach hereby developed.
```{r build-index, eval=run_safeguarded}
# check for missingness in data
rtrft_need_indtrs_fll %>% glimpse
  View()

# function for computing the relative index
rltv_indx<-function(x){
  x_mean <- mean(x, na.rm=T)
  rtio_mean <- x/x_mean
  return(rtio_mean)
}
  
rtrft_need_indx_v3_fll <-
rtrft_need_indtrs_fll %>%
  #98% winsorization
  datawizard::winsorize(
    threshold = 0.1, #or 0.5, for 95% winsorization
    method = "percentile",
    verbose = T
  ) %>%
  # weighting and weighted summation
  mutate(
    MentalHealth.p_rate.wt = MentalHealth.p_rate*wts_norm['wt5'],
    copd_asthma.p_rate.wt = copd_asthma.p_rate*wts_norm['wt1'],
    respiratory.p_rate.wt = respiratory.p_rate*wts_norm['wt2'],
    cvd.p_rate.wt = cvd.p_rate*wts_norm['wt3'],
    burns_falls_65_ova.p_rate.wt = burns_falls_65_ova.p_rate*wts_norm['wt4']
  ) %>% 
  rowwise() %>% 
  mutate(
    hlth_vlnblty_dmain = mean(c(MentalHealth.p_rate.wt, copd_asthma.p_rate.wt, respiratory.p_rate.wt, cvd.p_rate.wt, burns_falls_65_ova.p_rate.wt), na.rm = T),
  ) %>% 
  ungroup() %>% 
  select(-ends_with(c(".p_rate.wt", "_65_plus"))) %>% 
  mutate(
    # overall index: weighted product method
    rtrft_need_indx = epc_lss_eql_d_prcnt*income_score_rate*hlth_vlnblty_dmain,
    
    # Index (excluding health vulnerability: weighted product method
    rtrft_need_indx_mini = epc_lss_eql_d_prcnt*income_score_rate,
    
    # standardize domains and overall index
    across(
      c(epc_lss_eql_d_prcnt, income_score_rate, hlth_vlnblty_dmain, rtrft_need_indx, rtrft_need_indx_mini), .fns = list(z =datawizard::standardise, reltv_indx = rltv_indx),
      .names = "{.col}_{.fn}"
    ),
    
    # rank the index
    rtrft_need_indx_rnk = min_rank(desc(rtrft_need_indx)),
    rtrft_need_indx_mini_rnk = min_rank(desc(rtrft_need_indx_mini)),
    
    across(where(is.double), \(x) round(x, 4)),
  ) %>% 
  arrange(desc(rtrft_need_indx)) %>%
  glimpse

# rtrft_need_indx_v3 %>% 
#   glimpse
# 
# rtrft_need_indx_v3 %>% 
#   arrange(desc(rtrft_need_indx)) %>%
#   View
# 
# rtrft_need_indx_v3 %>% 
#   arrange(desc(rtrft_need_indx_mini)) %>%
#   View

# Export the declassified index. The condition-specific emergency-admission
# columns are removed before the file is written to the publication folder.
public_index <- rtrft_need_indx_v3_fll %>%
  select(-any_of(c(
    "asthma.p_rate",
    "respiratory.p_rate",
    "cvd.p_rate",
    "burns_falls_65_ova.p_rate"
  )))

write_csv(
  public_index,
  file.path(outputs_dir, "housing_retrofit_need_index_lsoa_v3.csv")
)
write_csv(public_index, public_index_path)

# # data prep for shiny app contd
# rtrft_need_indx_v3 %>%
#   select(LSOA11CD, LSOA11NM, LSOA21CD, LSOA21NM, hlth_vlnblty_dmain) %>% 
#   write.csv(file = paste0(tmp_fld,"/hlth_vlnblty_dmain.csv"), row.names = F)


```

```{r read-public-index}
# In a public run, this declassified file is the hand-off point past all
# safeguarded processing.
if (!file.exists(public_index_path)) {
  stop(
    "Missing ", public_index_path,
    ". This declassified input must be included in the publication repository."
  )
}

rtrft_need_indx_v3 <- read_csv(public_index_path, show_col_types = FALSE) %>% 
  mutate(across(
    all_of(c("epc_lss_eql_d_no", "rtrft_need_indx_rnk", "rtrft_need_indx_mini_rnk")), as.integer),
    across(where(is.character), factor)
    ) %>% 
  glimpse
```

# Map and plot the Index of housing retrofit need

## Import the shape file
Use data at LSOA and Local Authority District levels
Boundary data are retrieved at run time from: <https://geoportal.statistics.gov.uk/datasets/ons::lower-layer-super-output-areas-december-2011-boundaries-ew-bsc-v4/about>

<https://geoportal.statistics.gov.uk/datasets/ons::local-authority-districts-december-2011-boundaries-ew-bgc-2/about>
```{r geographic-boundaries}
# lsoa_21_sf <-
#   read_sf("open_data\\Lower_layer_Super_Output_Areas_December_2021_Boundaries_EW_BSC_V4_-5236167991066794441\\LSOA_2021_EW_BSC_V4.shp") %>% 
#   filter(str_starts(LSOA21CD, "E")) %>%
#   select(-c(LSOA21NMW, GlobalID)) %>%
#   glimpse

lsoa_11_sf <-
  read_sf(
    download_to_temp(source_urls$lsoa_2011_boundaries, ".geojson"),
    quiet = TRUE
  ) %>%
  clean_names() %>%
  rename(
    LSOA11CD = lsoa11cd,
    LSOA11NM = lsoa11nm,
    BNG_E = bng_e,
    BNG_N = bng_n,
    LONG = long,
    LAT = lat
  ) %>%
  filter(str_starts(LSOA11CD, "E")) %>%
  select(-any_of(c("lsoa11nmw", "global_id", "fid", "shape_area", "shape_length"))) %>%
  glimpse

# check the coordinate system
st_crs(lsoa_11_sf)

lad11_sf <-
  read_sf(
    download_to_temp(source_urls$lad_2011_boundaries, ".geojson"),
    quiet = TRUE
  ) %>%
  clean_names() %>%
  filter(str_starts(lad11cd, "E")) %>%
  select(-any_of(c("lad11nmw", "global_id", "fid", "shape_area", "shape_length"))) %>%
  glimpse

```

## Map the Index

Update: For better alignment with the data aggregation geographies utilised, and to mitigate missingness associated with misaligned aggregation zones, LSOA geographies have been reversed from 2021 to 2011

```{r map-index}
#spatialise the indicators
lsoa_sf_full <-
  lsoa_11_sf %>%
  select(-LSOA11NM) %>% #to avoid duplication
  left_join(rtrft_need_indx_v3, by = "LSOA11CD") %>%
  glimpse

# explore missingness
lsoa_sf_full %>% 
  st_drop_geometry() %>% 
  ExpData(2)

#possible colour schemes
col_schm = c(
  "-carto.tropic",
  "carto.army_rose",
  "-cols4all.pu_gn_div",
  "-powerbi.divergent",
  # "met.hokusai3", #-matplotlib.gn_bu, -kovesi.blue_cyan
  "-matplotlib.pi_yg",
  "hcl.green_orange",
  "kovesi.cy_gy_or",
  "kovesi.gn_wh_pu",
  # "-cols4all.pu_gn_div",
  "parks.olympic"
)

# cols4all::c4a_gui()

# the map objects and pictures
## Map by the Relative index
for (i in seq.int(1, 3)) {
  tictoc::tic()
  if (i == 1) {
    
    indx = c("epc_lss_eql_d_prcnt_reltv_indx",
             "income_score_rate_reltv_indx",
             "hlth_vlnblty_dmain_reltv_indx",
             "rtrft_need_indx_reltv_indx")
    sffx = c("Energy Inefficiency", "Income Deprivation", "Health Vulnerability", "Retrofit Needs")
    indtr_map_lst = list()
  }
  
  cat("\nMapping", indx[i], "...")
  
  indtr_map_lst[[indx[i]]] <-
    tm_shape(lsoa_sf_full) +
    tm_polygons(
      fill = indx[i],
      fill.scale = tm_scale_intervals(
        n = 7,
        values = col_schm[i], #can also vary the colour scheme here
        style = "fisher", #equal, pretty, jenks, fisher, quantile etc
        midpoint = 1
      ),
      fill.legend = tm_legend(
        title = paste0(sffx[i], "\n(Relative index)"),
        position = tm_pos_in("LEFT", "TOP"),
        item.r = 0,
        reverse = T,
        item.space = -0.2,
        item.na.space = 0,
        frame = F
      ),
      lwd = 0.0001,
      col_alpha = 0.0 #opacity 0 = 100% transparent
    ) +
    tm_shape(lad11_sf) +
    tm_borders(lwd=0.5) +
    tm_layout(frame.r = 0,
              panel.label.frame.r = 0
      # inner.margins = c(0.02, 0.02, 0.02, 0.2)
              )

  if(i ==max(seq.int(1, 3))){

    # use tmap_arrange to stitch maps of all 3 domains (on one row)
    indtr_map_dmains <-
    tmap_arrange(
      indtr_map_lst[1:3],
      nrow = 1,
      outer.margins = 0.01
    )

    # save stitched map
    cat("\nPlotting & saving mosaiced maps...")
    tmap_save(
      indtr_map_dmains,
      filename =  paste0(outputs_dir, "\\indtr_map_dmains_v4_fshr.png"),
      units = "cm",
      height = 15,
      width = 39,
      asp = 0
    )

    tictoc::toc()
  }
  # tictoc::toc()
}

# Facet index map of mini and full index
# Enable faceting for LAD maps
lad11_fcet_sf <- 
lad11_sf %>%
  mutate(rtrft_need_indx_z = NA, rtrft_need_indx_mini_z = NA) %>% 
  # filter(lad11nm == "Liverpool") %>% 
  pivot_longer(
    cols = c(rtrft_need_indx_z, rtrft_need_indx_mini_z),
    names_to = "Index_composition",
    values_to = "rtrft_need_indx_reltv",
    values_transform = as.double
  ) %>% #glimpse
  glimpse

# make map based on relative index

# # Conventional vs Enhanced index map (on thesame scale)
# indtr_map_fcet <-
#   lsoa_sf_full %>%
#   pivot_longer(
#   cols = c(rtrft_need_indx_mini_reltv_indx, rtrft_need_indx_reltv_indx),
#   names_to = "Index_composition",
#   values_to = "rtrft_need_indx_reltv",
#   values_transform = as.double
# ) %>% 
#     # glimpse
#   tm_shape() +
#   tm_polygons(
#     fill = "rtrft_need_indx_reltv",
#     fill.scale = tm_scale_intervals(
#       n = 10,
#       values = c(paletteer::paletteer_c("scico::vik", n = 10)), #was "-cols4all.pu_gn_div"
#       style = "fisher", #equal, pretty, fisher, jenks, quantile etc
#       midpoint = 1 #for z-score, change to 0
#     ),
#     fill.legend = tm_legend(
#       title = "Relative Index", # paste0(sffx[i]),
#       # title.align = "right",
#       position = tm_pos_out(pos.v = "top", pos.h = "right"),
#       item.r = 0,
#       reverse = T,
#       item.space = -0.2,
#       item.na.space = 0,
#       frame = F
#     ),
#     lwd = 0.0001,
#     col_alpha = 0.0  #opacity 0 = 100% transparent
#   ) +
#   # needs duplication to account for facets
#   tm_shape(lad11_fcet_sf) +
#   tm_borders(lwd=0.5) +
#   tm_facets_wrap(by = "Index_composition", ncol = 2) +
#   tm_layout(panel.labels = c("Conventional Index", "Enhanced Index"),
#             between_margin = 0.2,
#             frame.r = 0,
#             panel.label.frame.r = 0
#     # inner.margins = c(0.02, 0.02, 0.02, 0.2)
#             )
# 
# # save map as png file
# tmap_save(
#   indtr_map_fcet,
#   paste0(outputs_dir, "\\lsoa_need_relv_indx_maps_v4_fshr.png"),
#   units = "cm",
#   height = 13,
#   width = 26,
#   asp = 0,
#   # dpi = 600,
#   # scale = 0.85
# )

# Conventional vs Enhanced index map (on different scales)

indtr_map_fcet1 <-
  lsoa_sf_full %>%
  # mutate(reltv_indx_diff = rtrft_need_indx_reltv_indx - rtrft_need_indx_mini_reltv_indx) %>% 
  tm_shape() +
  tm_polygons(
    fill = c("rtrft_need_indx_mini_reltv_indx", "rtrft_need_indx_reltv_indx"),
    fill.scale = list(tm_scale_intervals(
      n = 10,
      values = "-cols4all.pu_gn_div", #was 
      style = "fisher", #equal, pretty, fisher, jenks, quantile etc
      midpoint = 1 #for z-score, change to 0
    ),
    tm_scale_intervals(
      n = 10,
      values = c(paletteer::paletteer_c("scico::vik", n = 10)), #"-cols4all.pu_gn_div"), #was 
      style = "fisher", #equal, pretty, fisher, jenks, quantile etc
      midpoint = 1 #for z-score, change to 0
    )
    ),
    # fill.free = T,
    fill.legend = list(tm_legend(
      title = "Relative Index", # paste0(sffx[i]),
      # title.align = "right",
      position = tm_pos_in(pos.v = "top", pos.h = "left"),
      item.r = 0,
      reverse = T,
      item.space = -0.2,
      item.na.space = 0,
      frame = F
    ),
    tm_legend(
      title = "Relative Index", # paste0(sffx[i]),
      # title.align = "right",
      position = tm_pos_in(pos.v = "top", pos.h = "left"),
      item.r = 0,
      reverse = T,
      item.space = -0.2,
      item.na.space = 0,
      frame = F
    )
    ),
    lwd = 0.0001,
    col_alpha = 0.0  #opacity 0 = 100% transparent
  ) +
  # needs duplication to account for facets
  tm_shape(lad11_fcet_sf) +
  tm_borders(lwd=0.5) +
  tm_layout(panel.labels = c("Conventional Index", "Enhanced Index"),
            between_margin = 0.2,
            frame.r = 0,
            panel.label.frame.r = 0
    # inner.margins = c(0.02, 0.02, 0.02, 0.2)
            )

# save map as png file
tmap_save(
  indtr_map_fcet1,
  paste0(outputs_dir, "\\lsoa_need_relv_indx1_maps_v4_fshr.png"),
  units = "cm",
  height = 13,
  width = 22,
  asp = 0,
  # dpi = 600,
  # scale = 0.85
)

# Enhanced index plus difference map

indtr_map_fcet2 <-
  lsoa_sf_full %>%
  mutate(reltv_indx_diff = rtrft_need_indx_reltv_indx - rtrft_need_indx_mini_reltv_indx) %>% 
  tm_shape() +
  tm_polygons(
    fill = c("rtrft_need_indx_reltv_indx", "reltv_indx_diff"),
    fill.scale = list(tm_scale_intervals(
      n = 10,
      values = c(paletteer::paletteer_c("scico::vik", n = 10)), #"-cols4all.pu_gn_div"), #was 
      style = "fisher", #equal, pretty, fisher, jenks, quantile etc
      midpoint = 1 #for z-score, change to 0
    ),
    tm_scale_intervals(
      n = 9,
      values = c(paletteer::paletteer_d("PrettyCols::PurpleTangerines", n = 9)), #"-cols4all.pu_gn_div"), #was 
      style = "fisher", #equal, pretty, fisher, jenks, quantile etc
      midpoint = 0 #for z-score, change to 0
    )
    ),
    # fill.free = T,
    fill.legend = list(tm_legend(
      title = "Relative Index", # paste0(sffx[i]),
      # title.align = "right",
      position = tm_pos_in(pos.v = "top", pos.h = "left"),
      item.r = 0,
      reverse = T,
      item.space = -0.2,
      item.na.space = 0,
      frame = F
    ),
    tm_legend(
      title = "Index Difference", # paste0(sffx[i]),
      # title.align = "right",
      position = tm_pos_in(pos.v = "top", pos.h = "left"),
      item.r = 0,
      reverse = T,
      item.space = -0.2,
      item.na.space = 0,
      frame = F
    )
    ),
    lwd = 0.0001,
    col_alpha = 0.0  #opacity 0 = 100% transparent
  ) +
  # needs duplication to account for facets
  tm_shape(lad11_fcet_sf) +
  tm_borders(lwd=0.5) +
  # tm_facets_wrap(by = "Index_composition", ncol = 2) +
  tm_layout(panel.labels = c("Conventional Index", "Difference: Enhanced - Conventional Index"),
            between_margin = 0.2,
            frame.r = 0,
            panel.label.frame.r = 0
    # inner.margins = c(0.02, 0.02, 0.02, 0.2)
            )

# save map as png file
tmap_save(
  indtr_map_fcet2,
  paste0(outputs_dir, "\\lsoa_need_relv_indx2_maps_v4_fshr.png"),
  units = "cm",
  height = 13,
  width = 22,
  asp = 0,
  # dpi = 600,
  # scale = 0.85
)

gc()
save.image()

```

# Statistical description & plots

## Boxplots
### Entire England
```{r}

# LSOA11toLAD20 contains 34,753 lsoas, while the lsoa21 contains 35,797 zones

rtrft_lsoa11_lad <- 
  lsoa_sf_full %>% 
  st_drop_geometry() %>% 
  select(-c(BNG_E, BNG_N, LAT, LONG)) %>%
  # glimpse
  mutate(
    LAD20NM = LAD20NM %>% factor() %>% fct_reorder(rtrft_need_indx_reltv_indx, .desc = F, .na_rm = T) 
  ) %>% 
  group_by(LAD20NM) %>% 
  # compute LAD median & mean
  mutate(
    rtrft_lad_mdn_indx = median(rtrft_need_indx, na.rm = T),
    rtrft_lad_reltv_indx_mean = mean(rtrft_need_indx_reltv_indx, na.rm = T)
  ) %>%
  ungroup() %>% 
  # arrange LAD by mean of index (SD)
  arrange(desc(rtrft_lad_mdn_indx), LAD20NM) %>%
  # for ploting
  mutate(
    lad_rnk = dense_rank(desc(LAD20NM)),
    lad_rnk2 = as.character(lad_rnk) %>% str_pad(width = 3, side = 'left', pad = 0),
    LAD20NM_rnk = paste0(LAD20NM, " [", lad_rnk2, "]"), 
    lad_facet = case_match(lad_rnk,
                           c(1:104)~1,
                           c(105:209)~2,
                           c(210:314)~3,
                           .default = lad_rnk),
    # tidyup
    lad_rnk2 = NULL
  ) %>% 
  glimpse

# rtrft_lsoa11_lad %>% 
#   glimpse

# 314 LAs
length(unique(rtrft_lsoa11_lad$LAD20CD))

# rtrft_lsoa11_lad %>% glimpse
# use boxplots to show variations in neighbourhood retrofit needs in local authorities (rtrft_need_indx_reltv_indx, rtrft_need_indx_z)
rtrft_lsoa11_lad_bxplt <-
  rtrft_lsoa11_lad %>% 
  ggplot(aes(y = forcats::fct_reorder(LAD20NM_rnk, rtrft_need_indx_reltv_indx, .desc = F, .na_rm = T), x = rtrft_need_indx_reltv_indx)) + #
  # for relative index, intercept = 1, but for z-score, intercept = 0
  geom_vline(xintercept = 1, linetype=6, colour = "black", linewidth = 0.6) + # A
  # geom_vline(xintercept = 0, linetype=6, colour = "black", linewidth = 0.6) +
  geom_boxplot(aes(fill = lad_rnk,
                   colour = lad_rnk,
                   group = LAD20NM_rnk),
               alpha = 0.5,
               outlier.size = 0.6,
               outlier.alpha = 0.3,
               linewidth = 0.2,
               show.legend = F
  )+
  scale_fill_gradient(low = "brown", high = "blue") +
  scale_color_gradient(low = "brown", high = "blue") +
  # coord_cartesian(xlim = c(-1,10)) + #or use NA
  # scale_x_continuous(breaks = scales::breaks_width(2), minor_breaks = NULL) +
  facet_wrap(vars(lad_facet), ncol= 3,
             scale = "free_y",
             labeller = as_labeller(c(`1`="1/3", `2`="2/3", `3`="3/3"))
  ) +
  labs(
    title = "Variations in Neighbourhood Need for Housing Retrofits in England's Local Authorities",
    # subtitle = "xxx",
    x = "Relative Index of Neighbourhood Need", # A
    # x = "Housing Retrofit Needs Index (z-score)", #B
    y = "Local Authority"
  ) +
  theme_linedraw() +
  theme(
    title = element_text(face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 10),
    axis.text.y = element_text(size = 6),
    panel.grid.major = element_line(linewidth = 0.05, colour = "grey40"),
    strip.text = element_text(face = "bold", size = 10)
  )

# ggplotly(rtrft_lsoa11_lad_bxplt)

ggsave(
  # plot b uses the index z-score, instead of the relative index
  filename = paste0(outputs_dir,"\\rtrft_lsoa11_lad_bxplt_v3a.png"),
  # filename = paste0(outputs_dir,"\\rtrft_lsoa11_lad_bxplt_v3b.png"),
  plot = rtrft_lsoa11_lad_bxplt,
  dpi = 600,
  units = "cm",
  height = 25,
  width = 35,
  # scale = 1.5,
  )
```

### Deciles of LAs (Top & bottom)
Because of the large number of LSOAs, we focus only on the top and bottom deciles in this plot.
```{r}

# Ensure the deciles of mini and full index is computed at the lsoa and median index values are available at the lad levels
rtrft_lsoa11_lad %<>% #glimpse
      mutate(
      mini_lsoa_indx_dcl = ntile(desc(rtrft_need_indx_mini), 10),
    fll_lsoa_indx_dcl = ntile(desc(rtrft_need_indx), 10)) %>% 
  group_by(LAD20CD) %>% 
    mutate(
    rtrft_lad_mdn_mini_indx = median(rtrft_need_indx_mini, na.rm = T)
    ) %>% 
  ungroup() %>% 
  glimpse

# strip to lad geographies and compute lad deciles
lad_indx_mdn_dcl <-
rtrft_lsoa11_lad %>%  
  dplyr::distinct(LAD20CD, .keep_all = T) %>% #glimpse
  mutate(
    cvnl_lad_indx_dcl = ntile(desc(rtrft_lad_mdn_mini_indx), 10),
    fll_lad_indx_dcl = ntile(desc(rtrft_lad_mdn_indx), 10)
         ) %>%
    select(LAD20CD, cvnl_lad_indx_dcl, fll_lad_indx_dcl) %>%
  glimpse

#   1st and 10th decides of LSOAs as boxplots at LAD level
rtrft_lsoa11_lad_bxplt_dcl <-
  rtrft_lsoa11_lad %>% 
  select(-c(starts_with(c("cvnl_lad_indx_dcl", "fll_lad_indx_dcl")))) %>%
  left_join(lad_indx_mdn_dcl, by = "LAD20CD") %>%
  filter(fll_lad_indx_dcl %in% c(1, 10)) %>% #glimpse
# frq(cvnl_lad_indx_dcl, fll_lad_indx_dcl)
  ggplot(aes(y = forcats::fct_reorder(LAD20NM_rnk, rtrft_need_indx_reltv_indx, .desc = F, .na_rm = T), x = rtrft_need_indx_reltv_indx)) + #
  # for relative index, intercept = 1, but for z-score, intercept = 0
  geom_vline(xintercept = 1, linetype=6, colour = "black", linewidth = 0.6) + # A
  # geom_vline(xintercept = 0, linetype=6, colour = "black", linewidth = 0.6) +
  geom_boxplot(aes(fill = lad_rnk,
                   colour = lad_rnk,
                   group = LAD20NM_rnk),
               alpha = 0.5,
               outlier.size = 0.6,
               outlier.alpha = 0.3,
               linewidth = 0.2,
               show.legend = F
  )+
  scale_fill_gradient(low = "brown", high = "blue") +
  scale_color_gradient(low = "brown", high = "blue") +
  # coord_cartesian(xlim = c(-1,10)) + #or use NA
  # scale_x_continuous(breaks = scales::breaks_width(2), minor_breaks = NULL) +
  facet_wrap(vars(fll_lad_indx_dcl), ncol= 2,
             scale = "free",
             labeller = as_labeller(c(`1`="Top decile", `10`="Bottom decile"))
  ) +
  labs(
    title = "Variations in Neighbourhood Need for Housing Retrofits in England's Local Authorities",
    # subtitle = "Top and bottom deciles of Need",
    x = "Relative Index of Neighbourhood Need", # A
    # x = "Housing Retrofit Needs Index (z-score)", #B
    y = "Local Authority"
  ) +
  theme_linedraw() +
  theme(
    title = element_text(face = "bold", size = 11),
    axis.title = element_text(face = "bold", size = 10),
    axis.text.y = element_text(size = 7),
    panel.grid.major = element_line(linewidth = 0.05, colour = "grey40"),
    strip.text = element_text(face = "bold", size = 10)
  )

ggsave(
  filename = paste0(outputs_dir,"\\rtrft_lsoa11_lad_bxplt_dcl_v4.png"),
  plot = rtrft_lsoa11_lad_bxplt_dcl,
  dpi = 600,
  units = "cm",
  height = 12,
  width = 25,
  # scale = 1.5,
  )

```

## Stats of index by LAs
Descriptive statistics of health-sensitive index values aggregated to the LA level. 
```{r}

indx_la_stats <-
  rtrft_lsoa11_lad %>% 
  centrality_description(x=LAD20NM,
                           y=rtrft_need_indx_reltv_indx,
                           type = "parametric") %>% 
  arrange(desc(rtrft_need_indx_reltv_indx)) %>% #glimpse
  rename(rtrft_need_indx_reltv_indx_mean = rtrft_need_indx_reltv_indx) %>% 
  select(-c(expression, n.expression)) %>% 
  mutate(
    across(where(is.numeric), \(x) round(x, 4))
  ) %>%
  glimpse

# export index at LA level
write_csv(indx_la_stats, paste0(outputs_dir, "\\housing_retrofit_need_index_LA_v3b.csv"))

```

# Associated inferential stats
Perform some complementary inferential statistics based on the foregoing boxplots of the indicators, such as ANOVA & student t tests.

## Heteroscedastic one-way ANOVA for trimmed means
```{r}
# One-way ANOVA (btw subjects design)
# ensure every LA has more than one LSOA to allow for parametric one-way ANOVA. This means dropping "Isles of Scilly" prior to computation, otherwise an error 'not enough observations' is returned.
rtrft_lsoa11_lad %>% 
  group_by(LAD20NM) %>% 
  summarise(lsoa_frq = n()) %>% 
  ungroup() %>% 
  arrange(lsoa_frq) %>% 
  glimpse

oneway_anova <- statsExpressions::oneway_anova(rtrft_lsoa11_lad %>% filter(LAD20NM != "Isles of Scilly"),
                           x=LAD20NM,
                           y=rtrft_need_indx_reltv_indx,
                           type = "robust") #"parametric"
oneway_anova

oneway_anova_df <- 
oneway_anova %>%
  select(-expression) %>% 
  mutate(across(where(is.numeric), \(x) round(x, 4))) %>% 
  pivot_longer(cols = everything(),
               names_to = "params",
               values_to = '1way_ANOVA',
               values_transform = as.character
               ) %>% 
    view("anova")

write.csv(oneway_anova_df, paste0(outputs_dir, "\\oneway_anova_df_v3.csv"))

# for reporting/interpretation purpose
aov(rtrft_need_indx_reltv_indx ~ LAD20NM, data = rtrft_lsoa11_lad %>% filter(LAD20NM != "Isles of Scilly")) %>% 
  report::report()
```
The ANOVA (formula: rtrft_need_indx_z ~ LAD20NM) suggests that:

  - The main effect of LAD20NM is statistically significant and large (F(312, 32530) = 46.37, p < .001; Eta2
= 0.31, 95% CI [0.30, 1.00])

Effect sizes were labelled following Field's (2013) recommendations.

## Pairwise comparison of the index between LAs
```{r}
# pairwise comparisons

pw_cmprn <- pairwise_comparisons(
  rtrft_lsoa11_lad %>% filter(LAD20NM != "Isles of Scilly"),
  x=LAD20NM,
  y=rtrft_need_indx_reltv_indx ,
  type = "parametric",
  var.equal = T
)

# of all 48,828 pairwise comparisons between LAs, 20,659 were statistically significant with p values of less than 0.05 as shown in the table below
la_pw_comp <- 
pw_cmprn %>%
  select(-expression) %>% 
  filter(p.value < 0.05) %>%
  mutate(
    across(where(is.numeric), \(x) round(x, 4))
  ) %>% 
  view("pairwise")
  # glimpse

# export pairwise comparisons of neighbourhood index of needs in LAs
write_csv(la_pw_comp, paste0(outputs_dir,"\\pairwise_t_test_of_need_index_LA_v3.csv"))

gc()
save.image()
```

# Enhanced index Correlation with the IMD (sub)domains
In this section, explore the bivariate association between the developed retrofit Index and other publicly available neighbourhood metrics to show that it is capturing a distinct aspect of place, relevant to population health and public policy.

To this end, correlation analysis between the index and other metrics will be performed.
Candidate metrics, include domains of the IMD, housing energy efficiency data, and indicators used to compute th energy deprivation classification.

```{r}
# data joins and var selections
rtft_indx_corr <- 
  rtrft_need_indx_v3 %>% 
  select(LSOA11CD, rtrft_need_indx, epc_lss_eql_d_prcnt) %>% 
  left_join(imd2019, by = c("LSOA11CD" = "lsoa_code_2011")) %>% 
  # drop some irrelevant IMD indicators
  select(-c(income_deprivation_affecting_children_index_idaci_score_rate, income_deprivation_affecting_older_people_idaopi_score_rate, children_and_young_people_sub_domain_score, where(is.character))) %>% 
  glimpse


#plot only significant correlations with Retrofit index

rtft_indx_corr_plt_v3 <-
  rtft_indx_corr %>% 
  # select(CIF_rsdl_1, all_of(obj3_covars)) %>%
  # rename(., any_of(nlp_rename_lookup)) %>%
  lares::corr_var(rtrft_need_indx,
           method = "pearson",
           top = 50,
           zeroes = F,
           ranks = T,
           # max_pvalue = 0.15, #significant at the 95% confidence interval
           save = F)

# extract the correlation table of the plots
rtft_indx_corr_df <-
  rtft_indx_corr_plt_v3$data %>%
  select(-c(pos, hjust)) %>% 
  mutate(across(where(is.double), \(x) round(x, 3)),
         variables = str_split_i(variables, " ", 2)
         ) %>% 
  view

```

# Sensitivity to health vulnerabiltiy
Show changes in retrofit needs prior to and after the inclusion of health vulnerability. Cleveland dot plots (to show difference in need value), heatmaps and sankey plots (to show changes in deciles) will be used, analysis can be performed at both LSOA and LA levels.

## All of the England
```{r}
# there are 32,844 LSOAs in england, so I show sensitivity of the top 50 to the inclusion of health vulnerability in the index

rtrft_need_indx_v3c <- 
rtrft_need_indx_v3 %>% 
  select(c(LSOA11NM, rtrft_need_indx_reltv_indx, rtrft_need_indx_mini_reltv_indx, ends_with("_rnk"))) %>%
  mutate(indx_reltv_indx_diff = rtrft_need_indx_mini_reltv_indx-rtrft_need_indx_reltv_indx,
         indx_rnk_diff = rtrft_need_indx_mini_rnk-rtrft_need_indx_rnk,
         indx_reltv_indx_diff_cat = if_else(rtrft_need_indx_reltv_indx < rtrft_need_indx_mini_reltv_indx, "Decrease", "Increase") %>% factor,
         indx_rnk_diff_cat = if_else(rtrft_need_indx_mini_rnk < rtrft_need_indx_rnk, "Decrease", "Increase") %>% factor,
         indx_reltv_indx_diff_abs = abs(indx_reltv_indx_diff),
         indx_rnk_diff_abs = abs(indx_rnk_diff)
         ) %>% 
  glimpse

# rtrft_need_indx_v3 %>% glimpse

indx_hlth_vuln_plt <- 
rtrft_need_indx_v3c %>% 
  slice_max(order_by = rtrft_need_indx_reltv_indx,
            n = 100) %>% 
  mutate(lsoa_facet = case_match(rtrft_need_indx_rnk,
                                c(1:50)~1,
                                c(51:100)~2,
                                .default = rtrft_need_indx_rnk)
         ) %>%
  ggplot(aes(x=reorder(LSOA11NM, rtrft_need_indx_reltv_indx))) +
  geom_segment(aes(xend=LSOA11NM,
                   # y=rtrft_need_indx_mini_reltv_indx,
                   # yend=rtrft_need_indx_reltv_indx,
                   
                    y = if_else(
                    indx_reltv_indx_diff_cat == "Increase",
                    rtrft_need_indx_mini_reltv_indx + 0.03,
                    rtrft_need_indx_mini_reltv_indx - 0.03
                  ),
                  yend = if_else(
                    indx_reltv_indx_diff_cat == "Increase",
                    rtrft_need_indx_reltv_indx - 0.04,
                    rtrft_need_indx_reltv_indx + 0.03
                  ),
                   
                   color=indx_reltv_indx_diff_cat),
               arrow = arrow(length = unit(0.2, "cm")),
               ) +
    # geom_point(aes(y=rtrft_need_indx_reltv_indx), color=rgb(0.2,0.7,0.1,1.0), size=5.5
    #          ) +
  geom_point(
    aes(
      y = rtrft_need_indx_reltv_indx,
      fill = "Health-sensitive"
    ),
    shape = 21,
    size = 6,
    alpha = 0.8,
    colour = "grey20",
    stroke = 0.3
  ) +
  geom_text(
    aes(y = rtrft_need_indx_reltv_indx, label = rtrft_need_indx_rnk),
    fontface = "bold",
    size = 3, color = "grey90"
  ) +
  # geom_point(aes(y=rtrft_need_indx_mini_reltv_indx), color=rgb(0.7,0.2,0.1,1.0), size=5.5) +
  geom_point(
    aes(
      y = rtrft_need_indx_mini_reltv_indx,
      fill = "Conventional"
    ),
    shape = 21,
    size = 6,
    alpha = 0.8,
    colour = "grey20",
    stroke = 0.3
  ) +
  geom_text(
    aes(y = rtrft_need_indx_mini_reltv_indx, label = rtrft_need_indx_mini_rnk),
    fontface = "bold",
    size = 2.5, color = "grey90"
  ) +
  scale_color_manual(
    name = "Change from conventional\nto health-sensitive index",
    values = c(
      "Increase" = rgb(0.7, 0.2, 0.1, 1.0),
      "Decrease" = rgb(0.2, 0.7, 0.1, 1.0)
    )
  ) +
  scale_fill_manual(
    name = "Index type",
    values = c(
      "Conventional" = rgb(0.7, 0.2, 0.1, 1.0),
      "Health-sensitive" = rgb(0.2, 0.7, 0.1, 1.0)
    )
  ) +
  coord_flip()+
  # guides(color =guide_legend(title = "Change from\nmini to full index")) +
  theme_linedraw() +
  facet_wrap(vars(lsoa_facet), ncol= 2,
             scale = "free",
             labeller = as_labeller(c(`1`="Rank 1 - 50", `2`="Rank 51 - 100"#
                                      # `3`="Third 1/3"
                                      ))
  ) +
  theme(
    # title = element_text(face = "bold", size = 10),
    # axis.title = element_text(face = "bold", size = 8),
    # axis.text.y = element_text(size = 6),
    panel.grid.major = element_line(linewidth = 0.05, colour = "grey40"),
    legend.position = "bottom"
  ) +
  labs(
    title = "Sensitivity of Retrofits Needs to Health Vulnerability",
    subtitle = "Top 100 Neighbourhoods with the Highest Need in England",
    y = "Relative Index of Need",
    x = "Neighbourhood (ordered by full index)",
    caption = "Labels of the dots represent neighbourhood ranks"
  )
indx_hlth_vuln_plt

# save plot
ggsave(
  filename = paste0(outputs_dir,"\\sntvty_indx_hlth_vuln_plt_v3.png"),
  plot = indx_hlth_vuln_plt,
  dpi = 600,
  units = "cm",
  height = 28,
  width = 30,
  # scale = 1.5,
  )

# notice that in the mini index, numerous (237) neighbourhodds are ranked 1. The is a consequence of the winsorization which capped extreme EPC and income deprivation scores to thesame values.


gc()
save.image()
```

## Update with LSOA to Region lookup
source <https://geoportal.statistics.gov.uk/datasets/ons::lsoa-2011-to-buasd-to-bua-to-lad-to-rgn-december-2011-best-fit-lookup-in-ew/about>
```{r region-lookup}
# LSOA to region mapping
lsoa_11_to_rgn <- 
  read_csv(source_urls$lsoa_2011_to_region, show_col_types = FALSE) %>%
  select(starts_with(c("LSOA11", "LAD11", "RGN11")), -ends_with(c("11NMW", "11NMW"))) %>%
  glimpse

# lsoa_11_to_rgn %>% glimpse
```

## Add regions, deciles and export data
THe is exported for use by the source r scipt for producing revised plots (heatmaps and sankey plots) below.
```{r}
rtrft_lsoa11_lad <- 
rtrft_lsoa11_lad %>% 
  select(-c(starts_with(c("RGN11NM", "RGN11CD")))) %>% 
  left_join(lsoa_11_to_rgn %>% select(-c(LSOA11NM, LAD11CD, LAD11NM)), by = "LSOA11CD") %>%
  select(-c(starts_with(c("cvnl_lad_indx_dcl", "fll_lad_indx_dcl")))) %>%
  left_join(lad_indx_mdn_dcl, by = "LAD20CD") %>%
  glimpse

# used to produce bolt-on plots of decile transitions
rtrft_lsoa11_lad %>% 
  write_csv(paste0(outputs_dir,"\\rtrft_lsoa11_lad_v4.csv"))
```

## Sensitivity plots: Heatmaps + Sankey plots
```{r}

# export bivariate correlation between the retrofits needs index and indicators of the IMD
write_csv(rtft_indx_corr_df %>% 
  mutate(across(where(is.double), \(x) round(x, 3))
         ),
  paste0(outputs_dir,"\\rtft_indx_imd_corr_df.csv")
  )

# create rtft_indx_imd_corr_df.csv and rtrft_lsoa11_lad_v4.csv before running the next line

source("scripts/create_retrofit_index_publication_plots_v2.R")
```


